---
title: "Data preparation"
author: "Julia"
date: "2025-08-12"
output:
  html_document: default
  pdf_document: default
---

## Brief comment

This markdown document was created to briefly show the outputs of the code included in the `dataset_preparation.R` as the full data will not be uploaded for confidentiality reasons. Therefore, this file will have the exact same code with small modifications to show that each step is doing.


### 1. Loading packages and baseline data sets

```{r}
# Loading packages

library(dplyr)  
library(tidyverse)

# Loading baseline data sets

alladults <- read.csv("C:/Users/julia/OneDrive - University of Edinburgh/EEB_MSc_UEd/Dissertation/diss/data/Adults_2025.csv")  # loading all adult blue tits data
as_tibble(alladults)  # visualising first rows of data
allbirdphen <- read.csv("C:/Users/julia/OneDrive - University of Edinburgh/EEB_MSc_UEd/Dissertation/diss/data/Bird_Phenology_2025.csv")  # loading all phenology data
as_tibble(allbirdphen)  # visualising first rows of data
nestlings <- read.csv("C:/Users/julia/OneDrive - University of Edinburgh/EEB_MSc_UEd/Dissertation/diss/data/NestlingsII.csv")  # loading all blue tit nestlings data
as_tibble(nestlings)  # visualising first rows of data
swaps <- read.csv("C:/Users/julia/OneDrive - University of Edinburgh/EEB_MSc_UEd/Dissertation/diss/data/clutchswaps.csv")
as_tibble(swaps)  # visualising first rows of data
```

### 2. Selecting variables that will be included in the final data sets 

Since we aim to study breeding success trends in blue tit populations across age groups and across sites, we will extract those variables that may be useful to study breeding success: first egg lay date (`fed`), clutch size ('cs`) and number of fledgelings (suc).

```{r}
# Selecting variables that will be useful for the project from the phenology data

blutiphen <- allbirdphen %>% filter(species == "bluti")  # first, we will select and extract cases that are of blue tits

blutiphen <- blutiphen %>% select(year, site, box, fed, cs, suc)  # we also select columns that will help us collate both databases into one by identifying individuals (i.e.: year, site and box)

head(blutiphen)

nrow(allbirdphen)

nrow(blutiphen)  # we end up with 2232 observations of breeding attempts

# Selecting variables that will be useful for the project from the adult data

adults <- alladults %>% filter(season != "winter", sex == "F")  # we will remove all data coming from adults captured in the winter (as we may not have their corresponding breeding season data)

nrow(alladults)

nrow(adults)  # after this filtering, we remove 2100 observations (of male birds and of winter recordings), and we have a total of 1828 recordings of female adults ringed and identified of which we should have breeding success data in blutiphen (or, at least, of most of them)

blutiadults <- adults %>% select(ring, year, site, box, age)  # finally, we select the columns that we will use

# Rearranging datasets in ascending order of years, site and nest boxes

level_order <- c("EDI", "RSY", "FOF", "BAD", "LVN", "DOW", "GLF", "SER", "MCH", "PTH", "STY", "BIR", "DUN", "BLG", "PIT", "KCK", "KCZ", "BLA", "CAL", "DNM", "DNC", "DNS", "DLW", "CRU", "NEW", "HWP", "INS", "FSH", "RTH", "AVI", "AVN", "CAR", "SLS", "TOM", "DAV", "ART", "MUN", "FOU", "ALN", "DEL", "TAI", "SPD", "OSP", "DOR")  # here I'm creating a variable that stores sites' codes in order of increasing latitude

blutiphen <- blutiphen %>% arrange(year, factor(site, levels = level_order), box) 

blutiadults <- blutiadults %>% arrange(year, factor(site, levels = level_order), box)  

```


### 3. Creating a new database by collating  blutiadults and blutiphen using columns "year", "site" and "box"

```{r}
# First, let's find out if there are duplicates by "year", "site" and "box" in both datasets:
blutiphen %>%
  count(year, site, box) %>%
  filter(n > 1)  # there seem to be 5 duplicates in blutiphen
```

These duplicates could be second broods (within the same breeding season and from the same female) or could be relays (if the first brood did not success):  
1. Case 2019, AVN site, box 2 seems to be a relay as the first entry has no success input.  
2. Case 2019, EDI site, box 1 is also this case (except for the fact that in the first entry there is a 0 instead of a NA).  
3. Case 2019, MUN site, box 1 also probably follows the re-lay case.  
4. And this is also probably the case 2021 site DOW, nestbox 4.

```{r}
blutiadults %>%
  count(year, site, box) %>%
  filter(n > 1)  # there seem to be 10 duplicates in blutiadults
```

These duplicates are re-measurements of the same females' tarsus length and body mass but later on (1-5 days later).

I'm particularly interested in removing duplicates so that they don't complicate calculating the age of each individual in each event we've recorded of them. I'm not particularly worried about which case is removed for `blutiadults` as both will store the same information. However, for `blutiphen` I am interested in storing the re-lays rather than the failed breeding attempts (ask Hannah confirmation about this). I'm going to try and rearrange my dataset so that relays appear before failed attempts as the `unique()` function in dplyr will keep the first row it encounters. I will keep the first rows because those are the ones for which most likely we have information on the females.

```{r}
blutiphen <- blutiphen %>% arrange(year, factor(site, levels = level_order), desc(suc), box)  # now, the first row in the duplicates should be the failed attempt

uqblutiphen <- distinct(blutiphen, year, site, box, .keep_all = T) 

head(uqblutiphen)

uqblutiadults <- distinct(blutiadults, ring, year, site, box, .keep_all = T)

head(uqblutiadults)
```

Now, not all adults are caught (e.g. when broods fail before nestlings reach they 10, when one of the parents dies or abandons the nest, etc.), therefore there might be observations in `blutiphen` that do not have a corresponding match in the `blutiadults` database. I will use the function `anti_join()` to find it out how many and which ones:

```{r}
no_adult_data <- uqblutiphen %>%
  anti_join(uqblutiadults, by = c("year", "site", "box"))  

nrow(no_adult_data)  
```

There are 412 rows in `blutiphen`for which we do not have data in `blutiadults`. I stored these problematic cases in their separate rows in case we want to reach for them in the future.

```{r}
no_bs_data <- uqblutiadults %>%
  anti_join(uqblutiphen, by = c("year", "site", "box"))  

nrow(no_bs_data)
```

There are also two adult female birds for which there is no matching breeding data. These are exceptional cases in which females were found incubating empty nests. These will be removed.

I will only include in the database the shared cases between blutiphen and blutiadults because lack of data of either of them is problematic.

```{r}
bluti1 <- uqblutiphen %>%
  inner_join(uqblutiadults, by =c("year", "site", "box")) 

head(bluti1)

nrow(bluti1)
```

There are some cases for which the number of fledgelings is -999. These are cases were nests were predated and, instead of indicating 0 as fledgeling success, -999 was written down in case they needed to be excluded at some point.

```{r}
# I will create a new bluti2 variable in which those cases show "NA" for fledgeling success

bluti2 <- bluti1

bluti2[which(bluti2$suc < 0),"suc"] <- NA
```

Additionally, I inspected cases where clutch size was greater than expected.

```{r}
nrow(bluti2[which(bluti2$cs > 14),])

# There are only three cases in the entire dataset where cs > 14. 

bluti2[which(bluti2$cs > 14),]

# Two of them belong to the same nest box and the same year, but have different females. 
# My guess was that two females were laying eggs in the same nest box, and since we cannot really tear appart to whom do both fed and suc correspond (and they're different ages as well), then the right call is to remove this cases entirely.
# As for the last one, I'd argue that is probably a very similar case except that the other female was never caught. I also removed the whole line (fed and suc included) for this reason:

bluti2 <- bluti2[-which(bluti2$cs > 14),]

bluti2 <- bluti2 %>% relocate(ring, year, site, box, age, fed, cs, suc)

nrow(bluti2)  # 1814 observations in total
```


### 4. Brief joint data set exploration

```{r}
unique(bluti2$year)  # we have data from years 2014-2025

bluti2 %>%
  count(year) %>%
  filter(n > 1)  # Number of cases/recordings per years

length(unique(bluti2$ring))  # We have recordings of a total of 1220 unique female birds across the years

nr_birds <- bluti2 %>%
  count(ring)  # storing how many times we have recorded each female

mean(nr_birds$n)  # On average, we have recorded 1.486885 breeding attempts of each female. 

# Most females only breed one year (or we only have one recording on average of each female; maybe they've had more but we haven't noticed)

max(nr_birds[, 2])  # the maximum number of breeding attempts recorded for the same female is 8...

nr_birds[which(nr_birds$n == max(nr_birds[, 2])),"ring"]  # ...and it's of female S921907

str(bluti2)  # checking that all columns are the right type

bluti2$suc <- as.numeric(bluti2$suc)  # changing number of fledgelings

# I think it would be interesting to add a column in which fledgeling success is a proportion (proportion of eggs laid that were successfully fledged). 
# This proportion will take the complete clutch into account, rather than the number of hatched eggs. However, it would be potentially more accurate to use the number of hatched eggs as base (maybe substracting the number of unhatched eggs found in V1 to the clutch size)

bluti2$suc_prop <- bluti2$suc/bluti2$cs

bluti2 <- bluti2 %>% relocate(suc_prop, .after=suc)

# Number of birds first recorded as yearlings

# Yearlings fall within age category 5, but in some cases people on the field mark them as age category 4 by mistake:

age4 <- bluti2 %>% filter(age == 4)  # there are  8 recordings of which age=4, which are most likely yearlings

age5 <- bluti2 %>% 
  filter(age <= 5) %>%
  group_by(ring) %>%
  count(ring)  # 765 unique female birds were first identified as yearlings

age6 <- bluti2 %>% 
  filter(age == 6) %>%
  group_by(ring) %>%
  count(ring)  # 697 unique females were captured as 2+ year olds

only_age6 <- anti_join(x = age6, y = age5)  # 559 females were first identified as 2+ year olds (~46% of all females)
```


### 5. Estimating birds' age and age at last reproduction

```{r}
nestlings <- nestlings %>%
  select(ring, year, site, box) %>%
  rename(ringedyear = year) %>%
  arrange(ringedyear, site, box)

bluti3 <- bluti2
bluti3$ringedyear <- NA
ring_nr <- 0
for (i in 1:nrow(bluti3)) {
  if (bluti3[i,"ring"] %in% nestlings$ring) {
    ring_nr <- as.character(bluti3[i,"ring"])
    row <- which(nestlings$ring == ring_nr)
    bluti3[i,"ringedyear"] <- nestlings[row, "ringedyear"]
  }
}  
length(which(!is.na(bluti3$ringedyear)))  
```

There are only `r length(which(!is.na(bluti3$ringedyear)))` (female) birds that hatched in the project and that were seen breeding in the nest boxes after that. 

We know exactly the hatching year of 30 female birds. For the remaining ones (N = 1448), we will need to estimate their age based on how we aged them on the field. We will assume that when we first get a bird in age category 6, the bird was 2 y.o. on that year:

```{r}
inds <- data.frame(ring = sort(unique(bluti3$ring)), 
                   firstcaught = tapply(bluti3$year, bluti3$ring, min), 
                   firstageclass = tapply(bluti3$age, bluti3$ring, min))  # defining a new dataframe with each unique female, the year she was first caught and the age she was 

inds$hatchyear <- ifelse(inds$firstageclass==6, (inds$firstcaught - 2), (inds$firstcaught - 1))  # calculating an estimated hatch year based on the age class when first caught (assuming age class 4 were yearlings)

bluti3$hatchyear <- inds$hatchyear[match(bluti3$ring, inds$ring)]  # matching hatch year back to original dataframe

for (i in 1:nrow(bluti3)) {
  if (!is.na(bluti3[i,"ringedyear"])) {
    bluti3[i, "hatchyear"] <- bluti3[i, "ringedyear"]
  }
}

bluti3$yo <- bluti3$year - bluti3$hatchyear  # calculating age

head(bluti3)

# Trying to create a "age at last reproduction" column

w <- tapply(bluti3$yo, bluti3$ring, max)

w <- as.data.frame(w)

w$rings <- rownames(w)

bluti3$w <- w$w[match(bluti3$ring, w$rings)]

# Rearranging the data set

blutidf <- bluti3 %>% 
  dplyr::select(ring, year, site, box, fed, cs, suc, hatchyear, ringedyear, age, yo, w) %>%
  dplyr::arrange(year, factor(site, levels = level_order), box)

```


### 6. Excluding number of fledgelings of nests involved in clutch swap experiments

Between 2017 and 2019, a clutch-swap experiment took place. The results of these experiments may have an influence on `suc`, and we should account for this. My aim is to inspect the data set to find the cases that took part in the experiment and switch their `suc` value into `NA,` as it may have been impacted by the experiment and, therefore, obscure age effects from the actual parents and of the surrounding environment.

```{r}
as_tibble(swaps)

swaps <- swaps %>% rename(box = origin.nest)  # re-naming to make things easier


# Assigning a case number to each case (row) in my database:
blutidf$case <- seq(from=1, to=nrow(blutidf), by=1)

# Now, I'll create a data frame that will store the cases in my database included in the clutch swap experiments
clutchswaps <- blutidf %>% 
  semi_join(swaps, by = c("year", "site", "box"))

clutchswaps %>% 
  count(year) # Number of cases involved in clutch swipe experiment per year
# According to this, 234 out of a total of 1660 were involved in the clutch swap experiment that took place between 2017 and 2019.

length(unique(clutchswaps$ring))  # It seems that a total of 201 female birds were involved in the experiment

# Now, based on the case number assigned to each row in the blutidf database, I will switch suc value to NA whenever the case is found in the dataframe that stores cases included in the experiment

for (i in 1:nrow(blutidf)) {
  if (blutidf[i,"case"] %in% clutchswaps$case) {
    blutidf[i,"suc"] <- NA
  }  
}


```


### 7. Creating separate data sets for recordings of females that are only between 1 and 3 years old for all three breeding traits

Although this section is included here, note that originally this cut was only done once exploration revealed 3 to be the age at which most chosen breeding traits peaked. For the sake of having in one single file all the steps to create the separate data sets (without overflowing it with plots), this code is included here.

```{r}
blutidf_3yo <- blutidf[which(blutidf$yo <= 3),]

length(unique(blutidf_3yo$ring))  

# We have 1219 females in total including 2025 data

population <- blutidf_3yo %>% 
  count(year,yo) %>%
  pivot_wider(names_from=year,values_from=c(n))

# There are many less 1-year-olds in 2025 than in any other year

#The number of observations of 3 years-old individuals per year is quite low overall, and it's almost always less than 20% of the population (2022 is the sole exception); birds aged 3 years old usually represent between 10 and 15% of the total of the (female) population.

mean(blutidf$w) 

 # females live on average 2.496141 years old on this study system
```

#### a) First egg lay date

```{r}
length(which(is.na(blutidf_3yo$fed)))
```

There are `r length(which(is.na(blutidf_3yo$fed)))` observations for which we have no first egg lay date recordings

```{r}
head(blutidf_3yo[which(is.na(blutidf_3yo$fed)),] ) 
```

Quite a few of these belong to sites that are visited less often (most sites in the Northern part of the transect, starting with HWP and until DOR, as well as sites in the southern transect that after a certain year started to receive visits less often: BAD, DLW and DUN) and, therefore, estimation of first egg lay date based on the number of eggs present in the nest box once its inspected is less accurate. However, and potentially more relevant, most of these fed-lacking observations belong to the year 2020, year in which the sanitary emergency caused by the COVID-19 pandemic provided an obstacle to conduct fieldwork following the standard protocol.

```{r}
fed_df <- blutidf_3yo %>% filter(!is.na(fed))

nrow(fed_df) # In the separate dataset for fed there are 1302 observations...

length(unique(fed_df$ring)) #... of 1006 unique (female) bird rings
```


#### b) Clutch size


```{r}
length(which(is.na(blutidf_3yo$cs)))
```

There are `r length(which(is.na(blutidf_3yo$cs)))` observations for which we have no clutch size recordings

```{r}
head(blutidf_3yo[which(is.na(blutidf_3yo$cs)),])  
```

As with `fed`, most of these clutch size-lacking observations belong to the year 2020.

```{r}
cs_df <- blutidf_3yo %>% filter(!is.na(cs))

nrow(cs_df) # In the separate dataset for cs there are 1600 observations...

length(unique(cs_df$ring)) #... of 1188 unique (female) bird rings
```


#### c) Number of fledgelings


```{r}
length(which(is.na(blutidf_3yo$suc)))
```

There are `r length(which(is.na(blutidf_3yo$suc)))` observations for which we have no clutch size recordings

```{r}
head(blutidf_3yo[which(is.na(blutidf_3yo$suc)),])  
```

As with `fed` and `suc`, most of these clutch size-lacking observations belong to the year 2020. However, these missing values also come from nests that were included in the swap experiments as well as from predated nests.

```{r}
suc_df <- blutidf_3yo %>% filter(!is.na(suc))

nrow(suc_df) # In the separate dataset for suc there are 1413 observations...

length(unique(suc_df$ring)) #... of 1079 unique (female) bird rings
```
