---
title: "Preparation of environmental predictors"
author: "Julia"
date: "2025-08-12"
output: html_document
---

## Brief comment

This markdown document was created to briefly show the outputs of the code included in the `environmental_predictors.R` file and in the `site_occupancy.R` file as the full data will not be uploaded for confidentiality reasons. Therefore, this file will have the exact same code with small modifications to show that each step is doing.


## 1. Loading packages and baseline data sets

```{r}
# Loading packages

library(dplyr)  
library(tidyverse)
library(tidyr)
library(vegan)

# Loading baseline data sets

blutidf <- read.csv("C:/Users/julia/OneDrive - University of Edinburgh/EEB_MSc_UEd/Dissertation/diss/data/blutidf_2025.csv")  # loading complete blue tit data set
blutidf_3yo <- read.csv("C:/Users/julia/OneDrive - University of Edinburgh/EEB_MSc_UEd/Dissertation/diss/data/blutidf_3yo_2025.csv")  # loading blue tit data set (observations limited by age <= 3)
sites <- read.csv("C:/Users/julia/OneDrive - University of Edinburgh/EEB_MSc_UEd/Dissertation/diss/data/site_detailsII.csv")  # loading site-specific data
habitat_foliage_scores <- read.csv("C:/Users/julia/OneDrive - University of Edinburgh/EEB_MSc_UEd/Dissertation/diss/data/Habitat_SiteII.csv")  # loading data on foliage score
habitat <- read.csv("C:/Users/julia/OneDrive - University of Edinburgh/EEB_MSc_UEd/Dissertation/diss/data/HabitatsII.csv")  # loading data used to calculate foliage score
allbirdphen <- read.csv("C:/Users/julia/OneDrive - University of Edinburgh/EEB_MSc_UEd/Dissertation/diss/data/Bird_Phenology_2025.csv")  # loading all phenology data
```

## 2. Creating a data frame that will store all environmental variables of potential interest:

```{r}
environment <- data.frame(sites = sites$site, longitude = sites$Mean.Long, latitude = sites$Mean.Lat, elevation = sites$Mean.Elev)

head(environment)
```

## 3. Adding variables (per site)

### a) Number of oak trees

```{r}
oak_sites <- habitat %>% select(site, nestbox, l_oak, m_oak, s_oak)

oak_sites <- oak_sites %>% replace_na(list(l_oak = 0, m_oak = 0, s_oak = 0))

oak_sites$TOTAL_oak <- 0

for (i in 1:nrow(oak_sites)) {
  oak_sites[i,"TOTAL_oak"] <- sum(c(oak_sites[i,"s_oak"], oak_sites[i,"m_oak"], oak_sites[i,"l_oak"]))
}

for (i in 1:nrow(environment)) {
  environment[i,"TOTAL_oak"] <- sum(oak_sites[which(oak_sites$site ==  environment[i,"sites"]),"TOTAL_oak"])
}
```

### b) Proportion of oak trees

```{r}
habitat_details <- habitat[,c(1,2,3,4)]

habitat <- habitat[,-c(1,2,3,4)]

habitat <- habitat %>% replace(is.na(.), 0)  # replacing all NAs for 0s

habitat <- cbind(habitat_details, habitat)

# Creating a dataframe summarising the number of trees per genus recorded at each nestbox, ignoring their girth
trees_per_box <- data.frame(sites = habitat$site, box = habitat$nestbox, alder = rep(0, nrow(habitat)), ash = rep(0, nrow(habitat)), aspen = rep(0, nrow(habitat)), beech = rep(0, nrow(habitat)), birch = rep(0, nrow(habitat)), blackthorn = rep(0, nrow(habitat)), cherry = rep(0, nrow(habitat)), chestnut = rep(0, nrow(habitat)), conifer = rep(0, nrow(habitat)), elder = rep(0, nrow(habitat)), elm = rep(0, nrow(habitat)), hawthorn = rep(0, nrow(habitat)), hazel = rep(0, nrow(habitat)), holly = rep(0, nrow(habitat)), juniper = rep(0, nrow(habitat)), lime = rep(0, nrow(habitat)), oak = rep(0, nrow(habitat)), pine = rep(0, nrow(habitat)), rose = rep(0, nrow(habitat)), rowan = rep(0, nrow(habitat)), sycamore = rep(0, nrow(habitat)), whitebeam = rep(0, nrow(habitat)), willow = rep(0, nrow(habitat)), yew = rep(0, nrow(habitat)), other = rep(0, nrow(habitat)))

trees_per_box$alder <- habitat$X6_alder + habitat$m_alder + habitat$s_alder  # calculating the number of each tree genus by adding recordings of the same tree genus of different girth
trees_per_box$ash <- habitat$l_ash + habitat$m_ash + habitat$s_ash
trees_per_box$aspen <- habitat$m_aspen + habitat$s_aspen
trees_per_box$beech <- habitat$l_beech + habitat$m_beech + habitat$s_beech + habitat$X6_beech
trees_per_box$birch <- habitat$l_birch + habitat$m_birch + habitat$s_birch + habitat$X21_birch + habitat$X6_birch
trees_per_box$blackthorn <- habitat$z_blackthorn + habitat$X21_blackthorn
trees_per_box$cherry <- habitat$l_cherry + habitat$m_cherry + habitat$s_cherry + habitat$z_cherry + habitat$X6_cherry
trees_per_box$chestnut <- habitat$m_chestnut + habitat$s_chestnut
trees_per_box$conifer <- habitat$l_conifer + habitat$m_conifer + habitat$s_conifer
trees_per_box$elder <- habitat$s_elder + habitat$X6_elder
trees_per_box$elm <- habitat$l_elm + habitat$m_elm + habitat$s_elm
trees_per_box$hawthorn <- habitat$s_hawthorn
trees_per_box$hazel <- habitat$s_hazel + habitat$X6_hazel + habitat$X21_hazel
trees_per_box$holly <- habitat$m_holly + habitat$s_holly + habitat$X6_holly + habitat$X21_holly
trees_per_box$juniper <- habitat$s_juniper + habitat$X6_juniper
trees_per_box$lime <- habitat$m_lime + habitat$s_lime
trees_per_box$oak <- habitat$l_oak + habitat$m_oak + habitat$s_oak
trees_per_box$pine <- habitat$l_pine + habitat$m_pine + habitat$s_pine
trees_per_box$rose <- habitat$X6_rose 
trees_per_box$rowan <- habitat$m_rowan + habitat$s_rowan + habitat$X6_rowan
trees_per_box$sycamore <- habitat$s_sycamore + habitat$l_sycamore + habitat$m_sycamore + habitat$X6_sycamore
trees_per_box$whitebeam <- habitat$m_whitebeam + habitat$s_whitebeam
trees_per_box$willow <- habitat$l_willow + habitat$m_willow + habitat$s_willow + habitat$z_willow + habitat$X6_willow + habitat$X21_willow
trees_per_box$yew <- habitat$m_yew + habitat$s_yew
trees_per_box$other <- habitat$s_other + habitat$z_other


# Calculating the total number of trees recorded at each nest box, irrespective of their genus
trees_per_box$total_trees <- trees_per_box$alder + trees_per_box$ash + trees_per_box$aspen + trees_per_box$beech + trees_per_box$birch + trees_per_box$blackthorn + trees_per_box$cherry + trees_per_box$chestnut + trees_per_box$conifer + trees_per_box$elder + trees_per_box$elm + trees_per_box$hawthorn + trees_per_box$hazel + trees_per_box$holly + trees_per_box$juniper + trees_per_box$lime + trees_per_box$oak + trees_per_box$pine + trees_per_box$rose + trees_per_box$rowan + trees_per_box$sycamore + trees_per_box$whitebeam + trees_per_box$willow + trees_per_box$yew + trees_per_box$other

# Creating a dataframe to store all trees per site rather than per nest box
trees <- data.frame(sites = sites$site, alder = rep(0, nrow(sites)), ash = rep(0, nrow(sites)), aspen = rep(0, nrow(sites)), beech = rep(0, nrow(sites)), birch = rep(0, nrow(sites)), blackthorn = rep(0, nrow(sites)), cherry = rep(0, nrow(sites)), chestnut = rep(0, nrow(sites)), conifer = rep(0, nrow(sites)), elder = rep(0, nrow(sites)), elm = rep(0, nrow(sites)), hawthorn = rep(0, nrow(sites)), hazel = rep(0, nrow(sites)), holly = rep(0, nrow(sites)), juniper = rep(0, nrow(sites)), lime = rep(0, nrow(sites)), oak = rep(0, nrow(sites)), pine = rep(0, nrow(sites)), rose = rep(0, nrow(sites)), rowan = rep(0, nrow(sites)), sycamore = rep(0, nrow(sites)), whitebeam = rep(0, nrow(sites)), willow = rep(0, nrow(sites)), yew = rep(0, nrow(sites)), other = rep(0, nrow(sites)))


for (i in 1:nrow(trees)) {
  trees[i,"alder"] <- sum(trees_per_box[which(trees_per_box$sites ==  trees[i,"sites"]),"alder"])
}  # adding all alders at all nest boxes per site

for (i in 1:nrow(trees)) {
  trees[i,"ash"] <- sum(trees_per_box[which(trees_per_box$sites ==  trees[i,"sites"]),"ash"])
}  # ash

for (i in 1:nrow(trees)) {
  trees[i,"aspen"] <- sum(trees_per_box[which(trees_per_box$sites ==  trees[i,"sites"]),"aspen"])
}  # aspen

for (i in 1:nrow(trees)) {
  trees[i,"beech"] <- sum(trees_per_box[which(trees_per_box$sites ==  trees[i,"sites"]),"beech"])
}  # beech

for (i in 1:nrow(trees)) {
  trees[i,"birch"] <- sum(trees_per_box[which(trees_per_box$sites ==  trees[i,"sites"]),"birch"])
}  # birch

for (i in 1:nrow(trees)) {
  trees[i,"cherry"] <- sum(trees_per_box[which(trees_per_box$sites ==  trees[i,"sites"]),"cherry"])
}  # cherry

for (i in 1:nrow(trees)) {
  trees[i,"blackthorn"] <- sum(trees_per_box[which(trees_per_box$sites ==  trees[i,"sites"]),"blackthorn"])
}  # blackthorn

for (i in 1:nrow(trees)) {
  trees[i,"chestnut"] <- sum(trees_per_box[which(trees_per_box$sites ==  trees[i,"sites"]),"chestnut"])
}  # chestnut

for (i in 1:nrow(trees)) {
  trees[i,"conifer"] <- sum(trees_per_box[which(trees_per_box$sites ==  trees[i,"sites"]),"conifer"])
}  # conifer

for (i in 1:nrow(trees)) {
  trees[i,"elder"] <- sum(trees_per_box[which(trees_per_box$sites ==  trees[i,"sites"]),"elder"])
}  # elder

for (i in 1:nrow(trees)) {
  trees[i,"elm"] <- sum(trees_per_box[which(trees_per_box$sites ==  trees[i,"sites"]),"elm"])
}  # elm

for (i in 1:nrow(trees)) {
  trees[i,"hawthorn"] <- sum(trees_per_box[which(trees_per_box$sites ==  trees[i,"sites"]),"hawthorn"])
}  # hawthorn

for (i in 1:nrow(trees)) {
  trees[i,"hazel"] <- sum(trees_per_box[which(trees_per_box$sites ==  trees[i,"sites"]),"hazel"])
}  # hazel

for (i in 1:nrow(trees)) {
  trees[i,"holly"] <- sum(trees_per_box[which(trees_per_box$sites ==  trees[i,"sites"]),"holly"])
}  # holly

for (i in 1:nrow(trees)) {
  trees[i,"juniper"] <- sum(trees_per_box[which(trees_per_box$sites ==  trees[i,"sites"]),"juniper"])
}  # juniper

for (i in 1:nrow(trees)) {
  trees[i,"lime"] <- sum(trees_per_box[which(trees_per_box$sites ==  trees[i,"sites"]),"lime"])
}  # lime

for (i in 1:nrow(trees)) {
  trees[i,"oak"] <- sum(trees_per_box[which(trees_per_box$sites ==  trees[i,"sites"]),"oak"])
}  # oak

for (i in 1:nrow(trees)) {
  trees[i,"pine"] <- sum(trees_per_box[which(trees_per_box$sites ==  trees[i,"sites"]),"pine"])
}  # pine

for (i in 1:nrow(trees)) {
  trees[i,"rose"] <- sum(trees_per_box[which(trees_per_box$sites ==  trees[i,"sites"]),"rose"])
}  # rose

for (i in 1:nrow(trees)) {
  trees[i,"rowan"] <- sum(trees_per_box[which(trees_per_box$sites ==  trees[i,"sites"]),"rowan"])
}  # rowan

for (i in 1:nrow(trees)) {
  trees[i,"sycamore"] <- sum(trees_per_box[which(trees_per_box$sites ==  trees[i,"sites"]),"sycamore"])
}  # sycamore

for (i in 1:nrow(trees)) {
  trees[i,"whitebeam"] <- sum(trees_per_box[which(trees_per_box$sites ==  trees[i,"sites"]),"whitebeam"])
}  # whitebeam

for (i in 1:nrow(trees)) {
  trees[i,"willow"] <- sum(trees_per_box[which(trees_per_box$sites ==  trees[i,"sites"]),"willow"])
}  # willow

for (i in 1:nrow(trees)) {
  trees[i,"yew"] <- sum(trees_per_box[which(trees_per_box$sites ==  trees[i,"sites"]),"yew"])
}  # yew

for (i in 1:nrow(trees)) {
  trees[i,"other"] <- sum(trees_per_box[which(trees_per_box$sites ==  trees[i,"sites"]),"other"])
}  # other


trees$total_trees <- trees$alder + trees$ash + trees$aspen + trees$beech + trees$birch + trees$blackthorn + trees$cherry + trees$chestnut + trees$conifer + trees$elder + trees$elm + trees$hawthorn + trees$hazel + trees$holly + trees$juniper +trees$lime + trees$oak + trees$pine + trees$rose + trees$rowan + trees$sycamore + trees$whitebeam + trees$willow + trees$yew + trees$other 

trees$oak_proportion <- trees$oak/trees$total_trees

environment$proportion_oak <- trees$oak_proportion
```


### c) Number of oak and birch trees combined

```{r}
trees$oak_birch <- trees$birch + trees$oak

environment$oak_birch <- trees$oak_birch
```


### d) Proportion of oak and birch trees

```{r}
trees$proportion_oak_birch <- trees$oak_birch/trees$total_trees

environment$proportion_oak_birch <- trees$proportion_oak_birch
```


### e) Number of oak, birch and sycamore trees combined

```{r}
trees$oak_birch_sycamore <- trees$birch + trees$oak + trees$sycamore

environment$oak_birch_sycamore <- trees$oak_birch_sycamore
```


### f) Proportion of oak, birch and sycamore trees combined

```{r}
trees$proportion_oak_birch_sycamore <- trees$oak_birch_sycamore/trees$total_trees

environment$proportion_oak_birch_sycamore <- trees$proportion_oak_birch_sycamore
```


### g) Tree diversity (at genus level)

```{r}
trees <- trees %>% relocate(c(conifer, other), .after = yew)

# I will be using the Simpson's Index to calculate tree diversity, although I could use other index. Need to justify why one or the other

tree.simp <- diversity(trees[,2:24],index = "simpson")  # I'm excluding the columns "other" and "conifers" because I don't exactly know what taxa they comprise and whether said taxa (or some of it) are already present in other columns (f.e. pine trees)

environment$tree_diversity_simpson <- tree.simp
```


### h) Total foliage score

```{r, fig.align='center'}
level_order <- c("EDI", "RSY", "FOF", "BAD", "LVN", "DOW", "GLF", "SER", "MCH", "PTH", "STY", "BIR", "DUN", "BLG", "PIT", "KCK", "KCZ", "BLA", "CAL", "DNM", "DNC", "DNS", "DLW", "CRU", "NEW", "HWP", "INS", "FSH", "RTH", "AVI", "AVN", "CAR", "SLS", "TOM", "DAV", "ART", "MUN", "FOU", "ALN", "DEL", "TAI", "SPD", "OSP", "DOR")
habitat_foliage_scores <- habitat_foliage_scores %>% arrange(factor(Site, levels = level_order))

fs <- habitat_foliage_scores[,1:12]  # selecting columns with absolute foliage scores

score <- fs %>% pivot_longer(cols=c(Alder_FS, Ash_FS, Aspen_FS, Beech_FS, Birch_FS, Elm_FS, Oak_FS, Sycamore_FS, Willow_FS, Conifer_FS, OthDecid_FS), names_to = "foliage_score")

ggplot(score, aes(x=Site, y=value, fill = foliage_score)) +
  geom_col(colour="black") +
  theme_bw() +
  labs(x = "Sites", y = "Foliage scores") +
  guides(fill=guide_legend(title="Foliage score")) +
  scale_fill_hue(labels = c("Alder", "Ash", "Aspen", "Beech", "Birch", "Conifer", "Elm", "Oak", "Other deciduous", "Sycamore", "Willow")) +
  theme(axis.title = element_text(size = 13), 
        axis.text = element_text(size = 11),
        axis.text.x = element_text(angle = 90, margin = unit(c(3, 0, 0, 0), "mm")), 
        axis.title.x = element_text(margin = unit(c(3, 0, 0, 0), "mm")),
        axis.title.y = element_text(margin = unit(c(0, 5, 0, 0), "mm")))

environment$total_FS <- habitat_foliage_scores[match(environment$sites, habitat_foliage_scores$Site), "Total"]
```


### i) Oak foliage score

```{r}
environment$oak_FS <- habitat_foliage_scores[match(environment$sites, habitat_foliage_scores$Site),"Oak_FS"]
```

### j) Birch foliage score

```{r}
environment$birch_FS <- habitat_foliage_scores[match(environment$sites, habitat_foliage_scores$Site),"Birch_FS"]
```


### k) Sycamore foliage score

```{r}
environment$sycamore_FS <- habitat_foliage_scores[match(environment$sites, habitat_foliage_scores$Site),"Sycamore_FS"]
```


### l) Willow foliage score

```{r}
environment$willow_FS <- habitat_foliage_scores[match(environment$sites, habitat_foliage_scores$Site),"Willow_FS"]
```


### m) Beech foliage score

```{r}
environment$beech_FS <- habitat_foliage_scores[match(environment$sites, habitat_foliage_scores$Site),"Beech_FS"]
```


### n) Oak, birch and sycamore foliage score

```{r}
oak_birch_sycamore_FS <- habitat_foliage_scores %>% 
  select(c(Site,Oak_FS,Birch_FS,Sycamore_FS)) %>%
  mutate(sum = Oak_FS + Birch_FS + Sycamore_FS)

environment$oak_birch_sycamore_FS <- oak_birch_sycamore_FS[match(environment$sites, oak_birch_sycamore_FS$Site),"sum"]
```


### o) Oak and birch foliage score

```{r}
oak_birch_FS <- habitat_foliage_scores %>% 
  select(c(Site,Oak_FS,Birch_FS)) %>%
  mutate(sum = Oak_FS + Birch_FS)

environment$oak_birch_FS <- oak_birch_FS[match(environment$sites, oak_birch_FS$Site),"sum"]
```


### p) Oak foliage score (proportion)

```{r}
environment$oak_FS_prop <- habitat_foliage_scores[match(environment$sites, habitat_foliage_scores$Site), "Oak_prop"]
```


### q) Oak, sycamore and birch foliage score (proportion)

```{r}
oak_birch_sycamore_FS_prop <- habitat_foliage_scores %>% 
  select(c(Site,Oak_prop,Birch_prop,Sycamore_prop)) %>%
  mutate(sum = Oak_prop + Birch_prop + Sycamore_prop)

environment$oak_birch_sycamore_FS_prop <- oak_birch_sycamore_FS_prop[match(environment$sites, oak_birch_sycamore_FS_prop$Site),"sum"]
```


### r) Oak and birch foliage score (proportion)

```{r}
oak_birch_FS_prop <- habitat_foliage_scores %>% 
  select(c(Site,Oak_prop,Birch_prop,Sycamore_prop)) %>%
  mutate(sum = Oak_prop + Birch_prop)

environment$oak_birch_FS_prop <- oak_birch_FS_prop[match(environment$sites, oak_birch_FS_prop$Site),"sum"]
```


### s) Deciduous trees foliage score:

```{r}
environment$alder_FS <- habitat_foliage_scores[match(environment$sites, habitat_foliage_scores$Site),"Alder_FS"]
environment$ash_FS <- habitat_foliage_scores[match(environment$sites, habitat_foliage_scores$Site),"Ash_FS"]
environment$aspen_FS <- habitat_foliage_scores[match(environment$sites, habitat_foliage_scores$Site),"Aspen_FS"]
environment$elm_FS <- habitat_foliage_scores[match(environment$sites, habitat_foliage_scores$Site),"Elm_FS"]
environment$OthDecid_FS <- habitat_foliage_scores[match(environment$sites, habitat_foliage_scores$Site),"OthDecid_FS"]

environment <- environment %>% 
  mutate(all_decid_FS = alder_FS + ash_FS + aspen_FS + beech_FS + birch_FS + elm_FS + oak_FS + sycamore_FS + willow_FS + OthDecid_FS)
```


### t) Site occupancy

```{r}
birdphen <- allbirdphen[-which(allbirdphen$X2brood.relay == 1),]  # removing relays because they would artificially increase the number of nest boxes
occupancy <- data.frame(site = rep(c("EDI", "RSY", "FOF", "BAD", "LVN", "DOW", "GLF", "SER", "MCH", "PTH", "STY", "BIR", "DUN", "BLG", "PIT", "KCK", "KCZ", "BLA", "CAL", "DNM", "DNC", "DNS", "DLW", "CRU", "NEW", "HWP", "INS", "FSH", "RTH", "AVI", "AVN", "CAR", "SLS", "TOM", "DAV", "ART", "MUN", "FOU", "ALN", "DEL", "TAI", "SPD", "OSP", "DOR"), each = 11), year = rep(2014:2024, 44), total_boxes = 0, occ_boxes = 0)

for (i in 1:nrow(birdphen)) {
  for (a in 1:nrow(occupancy)) {
    if ((occupancy[a,"site"] == birdphen[i, "site"]) & (occupancy[a,"year"] == birdphen[i, "year"])) {
      occupancy[a, "total_boxes"] <- occupancy[a, "total_boxes"] + 1
    }
  }
}

for (i in 1:nrow(birdphen)) {
  for (a in 1:nrow(occupancy)) {
    if ((occupancy[a,"year"] == birdphen[i, "year"]) & (occupancy[a,"site"] == birdphen[i, "site"])) {
      if ((!is.na(birdphen[i,"fki"])) & (birdphen[i, "species"] == "bluti")) {
        occupancy[a, "occ_boxes"] <- occupancy[a, "occ_boxes"] + 1
      }
    }
  }
}

for (i in 1:nrow(occupancy)) {
  if (occupancy[i, "total_boxes"] == 0) {
    occupancy[i, "occ_boxes"] <- NA
  }
}

occupancy$prop_occ <- occupancy$occ_boxes/occupancy$total_boxes


environment[which(environment$sites =="EDI"), "site_occ"] <- mean(na.omit(occupancy[which(occupancy$site == "EDI"), "prop_occ"]))  

environment[which(environment$sites =="RSY"), "site_occ"] <- mean(na.omit(occupancy[which(occupancy$site == "RSY"), "prop_occ"]))  

environment[which(environment$sites =="FOF"), "site_occ"] <- mean(na.omit(occupancy[which(occupancy$site == "FOF"), "prop_occ"]))  

environment[which(environment$sites =="BAD"), "site_occ"] <- mean(na.omit(occupancy[which(occupancy$site == "BAD"), "prop_occ"]))  

environment[which(environment$sites =="LVN"), "site_occ"] <- mean(na.omit(occupancy[which(occupancy$site == "LVN"), "prop_occ"]))  

environment[which(environment$sites =="DOW"), "site_occ"] <- mean(na.omit(occupancy[which(occupancy$site == "DOW"), "prop_occ"]))  

environment[which(environment$sites =="GLF"), "site_occ"] <- mean(na.omit(occupancy[which(occupancy$site == "GLF"), "prop_occ"]))  

environment[which(environment$sites =="SER"), "site_occ"] <- mean(na.omit(occupancy[which(occupancy$site == "SER"), "prop_occ"]))  

environment[which(environment$sites =="MCH"), "site_occ"] <- mean(na.omit(occupancy[which(occupancy$site == "MCH"), "prop_occ"]))  

environment[which(environment$sites =="PTH"), "site_occ"] <- mean(na.omit(occupancy[which(occupancy$site == "PTH"), "prop_occ"]))  

environment[which(environment$sites =="STY"), "site_occ"] <- mean(na.omit(occupancy[which(occupancy$site == "STY"), "prop_occ"]))  

environment[which(environment$sites =="BIR"), "site_occ"] <- mean(na.omit(occupancy[which(occupancy$site == "BIR"), "prop_occ"]))  

environment[which(environment$sites =="DUN"), "site_occ"] <- mean(na.omit(occupancy[which(occupancy$site == "DUN"), "prop_occ"]))  

environment[which(environment$sites =="BLG"), "site_occ"] <- mean(na.omit(occupancy[which(occupancy$site == "BLG"), "prop_occ"]))  

environment[which(environment$sites =="PIT"), "site_occ"] <- mean(na.omit(occupancy[which(occupancy$site == "PIT"), "prop_occ"]))  

environment[which(environment$sites =="KCK"), "site_occ"] <- mean(na.omit(occupancy[which(occupancy$site == "KCK"), "prop_occ"]))  

environment[which(environment$sites =="KCZ"), "site_occ"] <- mean(na.omit(occupancy[which(occupancy$site == "KCZ"), "prop_occ"]))  

environment[which(environment$sites =="BLA"), "site_occ"] <- mean(na.omit(occupancy[which(occupancy$site == "BLA"), "prop_occ"]))  

environment[which(environment$sites =="CAL"), "site_occ"] <- mean(na.omit(occupancy[which(occupancy$site == "CAL"), "prop_occ"]))  

environment[which(environment$sites =="DNM"), "site_occ"] <- mean(na.omit(occupancy[which(occupancy$site == "DNM"), "prop_occ"]))  

environment[which(environment$sites =="DNC"), "site_occ"] <- mean(na.omit(occupancy[which(occupancy$site == "DNC"), "prop_occ"]))  

environment[which(environment$sites =="DNS"), "site_occ"] <- mean(na.omit(occupancy[which(occupancy$site == "DNS"), "prop_occ"]))  

environment[which(environment$sites =="DLW"), "site_occ"] <- mean(na.omit(occupancy[which(occupancy$site == "DLW"), "prop_occ"]))  

environment[which(environment$sites =="CRU"), "site_occ"] <- mean(na.omit(occupancy[which(occupancy$site == "CRU"), "prop_occ"]))  

environment[which(environment$sites =="NEW"), "site_occ"] <- mean(na.omit(occupancy[which(occupancy$site == "NEW"), "prop_occ"]))  

environment[which(environment$sites =="HWP"), "site_occ"] <- mean(na.omit(occupancy[which(occupancy$site == "HWP"), "prop_occ"]))  

environment[which(environment$sites =="INS"), "site_occ"] <- mean(na.omit(occupancy[which(occupancy$site == "INS"), "prop_occ"]))  

environment[which(environment$sites =="FSH"), "site_occ"] <- mean(na.omit(occupancy[which(occupancy$site == "FSH"), "prop_occ"]))  

environment[which(environment$sites =="RTH"), "site_occ"] <- mean(na.omit(occupancy[which(occupancy$site == "RTH"), "prop_occ"]))  

environment[which(environment$sites =="AVI"), "site_occ"] <- mean(na.omit(occupancy[which(occupancy$site == "AVI"), "prop_occ"]))  

environment[which(environment$sites =="AVN"), "site_occ"] <- mean(na.omit(occupancy[which(occupancy$site == "AVN"), "prop_occ"]))  

environment[which(environment$sites =="CAR"), "site_occ"] <- mean(na.omit(occupancy[which(occupancy$site == "CAR"), "prop_occ"]))  

environment[which(environment$sites =="SLS"), "site_occ"] <- mean(na.omit(occupancy[which(occupancy$site == "SLS"), "prop_occ"]))  

environment[which(environment$sites =="TOM"), "site_occ"] <- mean(na.omit(occupancy[which(occupancy$site == "TOM"), "prop_occ"]))  

environment[which(environment$sites =="DAV"), "site_occ"] <- mean(na.omit(occupancy[which(occupancy$site == "DAV"), "prop_occ"]))  

environment[which(environment$sites =="ART"), "site_occ"] <- mean(na.omit(occupancy[which(occupancy$site == "ART"), "prop_occ"]))  

environment[which(environment$sites =="MUN"), "site_occ"] <- mean(na.omit(occupancy[which(occupancy$site == "MUN"), "prop_occ"]))  

environment[which(environment$sites =="FOU"), "site_occ"] <- mean(na.omit(occupancy[which(occupancy$site == "FOU"), "prop_occ"]))  

environment[which(environment$sites =="ALN"), "site_occ"] <- mean(na.omit(occupancy[which(occupancy$site == "ALN"), "prop_occ"]))  

environment[which(environment$sites =="DEL"), "site_occ"] <- mean(na.omit(occupancy[which(occupancy$site == "DEL"), "prop_occ"]))  

environment[which(environment$sites =="TAI"), "site_occ"] <- mean(na.omit(occupancy[which(occupancy$site == "TAI"), "prop_occ"]))  

environment[which(environment$sites =="SPD"), "site_occ"] <- mean(na.omit(occupancy[which(occupancy$site == "SPD"), "prop_occ"]))  

environment[which(environment$sites =="OSP"), "site_occ"] <- mean(na.omit(occupancy[which(occupancy$site == "OSP"), "prop_occ"]))  

environment[which(environment$sites =="DOR"), "site_occ"] <- mean(na.omit(occupancy[which(occupancy$site == "DOR"), "prop_occ"]))  

environment <- environment[,-5]
```


## 3. Final environmental predictors data set

```{r}
head(environment)
```

